The title

### 2016-10-04 analysis/paper/paper.Rmd
more on intro

The Irrawaddy River terraces of central Myanmar have been part of global Pleistocene archaeology since de Terra and Movius [-@deTerra1943] published their monograph reporting archaeological evidence from the Middle Pleistocene. Their expedition also resulted in Movius' [-@movius1948lower] influential paper claiming to identify a major cultural and technological division between the east and west hemispheres. Movius described a region to the west of the Irrawaddy where hand-axes are present in Middle Pleistocene contexts, and a region to the east where hand-axes are absent, with chopper-chopping tools found instead. This division has subsequently been known as the Movius Line, and has been the subject of extensive discussion by archaeologists [@lycett2010demographic; @petraglia2008large; @lycett2010movius; keates2002movius; @norton2006movius; @yamei2000mid; @norton2008movius; @brantingham1999astride; @brumm2012biface; @brumm2010movius; @norton2006middle; @davis1999recent; @norton2000current; @lycett2007soanian; Romanowska2016]. This includes critical discussion that has questioned the reliability of Movius' observations

The importance of the Movius' findings depend on the chronology of the artefacts he discovered.  inferred from his geomorphological interpretations. Despite the prominence of the Movius Line in discussions of global patterns of human evolution, there has not been any new data from the location of Line since Movius' initial field work. In this paper we assess the current status of the

@@

Descriptions of the physical and cultural geography of the Irrawaddy terraces of central Myanmar have been appearing scholarly literature since Myanmar was attached to the British Empire in 1886. In the following section we review this literature to characterise this landscape. Then we review previous work on the archaeology of this landscape. Finally, we report on our recent surveys in the Chauk region, and discuss the implications for the archaeological significance of the region.

@@

# Dennell is a major commentator recently
Dennell is one of the more recent critics of the concept of the Movius Line. In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life]. The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution. According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene. Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
In contrast to Dennell’s views, Norton et al. [-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line. The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].
Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life]. Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
The use of stone tools may have been influenced by certain situations, tasks or individuals. Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times. Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].

>>>>

The Irrawaddy River terraces of central Myanmar have been part of global Pleistocene archaeology since de Terra and Movius [-@deTerra1943] published their monograph reporting archaeological evidence from the Middle Pleistocene. Their expedition also resulted in Movius' [-@movius1948lower] influential paper claiming to identify a major cultural and technological division between the east and west hemispheres. Movius described a region to the west of the Irrawaddy where hand-axes are present in Middle Pleistocene contexts, and a region to the east where hand-axes are absent, with chopper-chopping tools found instead. This division has subsequently been known as the Movius Line, and has been the subject of extensive discussion by archaeologists [@lycett2010demographic; @petraglia2008large; @lycett2010movius; keates2002movius; @norton2006movius; @yamei2000mid; @norton2008movius; @brantingham1999astride; @brumm2012biface; @brumm2010movius; @norton2006middle; @davis1999recent; @norton2000current; @lycett2007soanian; Romanowska2016]. This includes critical discussion that has questioned the reliability of Movius' field observations [@dennell2016life; @dennell2014hallam].

The importance of the Movius' findings depend primarily on the chronology of the landscape he described, and the technology of the artefacts that he collected.  Despite the prominence of the Movius Line in discussions of global patterns of human evolution, there has not been any new data from the location of Line since Movius' initial field work. Movius' interpretation of the Irrawaddy terraces and the technology of the artefacts he described remains untested by new data from his study area. In this paper we assess the current status of the Irrawaddy terraces and stratigraphy, and the technology of the artefacts found near the Irrawaddy in central Myanmar. We report on our recent surveys and field observations of the key archaeological locations that Movius visited in January of 1938.

@@

Descriptions of the physical and cultural geography of the Irrawaddy terraces of central Myanmar began appearing with increasing frequency in scholarly literature after Myanmar was attached to the British Empire in 1886. The Japanese occupation in 1942 disrupted geological research, and few publications appeared after the Second World War [@bender1983geology].  In the following sections we review this literature to characterise this landscape. Then we review previous work on the archaeology of this landscape.

@@

# Dennell is a major commentator recently
Dennell is one of the more recent critics of the concept of the Movius Line. In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life]. The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution. According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene. Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].

In contrast to Dennell’s views, Norton et al. [-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line. The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].

Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life]. Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.

The use of stone tools may have been influenced by certain situations, tasks or individuals. Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times. Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].
